---
title: "Reproducible Research - Week 4 Peer Project"
author: "Vignesh C Iyer"
date: "7/14/2020"
output:
  html_document:
    keep_md: yes
---

## Synopsis

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

## Assignment
The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events. You must use the database to answer the questions below and show the code for your entire analysis. Your analysis can consist of tables, figures, or other summaries. You may use any R package you want to support your analysis.

## Data

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the course web site:

[Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

* National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

* National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

### Data Pre-processing

The Storm Data is fetched, downloaded to the local system and then its contents are read based on the code given below

```{r}
# This section deals with the downloading the compressed file and
# extracting it contents.

stormData <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

# The file is downloaded using the download.file function.
download.file(stormData, destfile = "../StormData.csv.bz2")

# reading data from the file
readStormData <- read.csv("../StormData.csv.bz2")

# Fetching column names of Storm Data using the colNames function
colnames(readStormData)

str(readStormData)

# Fetching first few rows of Storm Data
head(readStormData)

# fetching the unique event type in the Storm Data
head(unique(readStormData$EVTYPE))

```

We notice that the Date format is that of a Character from the below code

```{r}
class(readStormData$BGN_DATE)
```

We will convert it to Date format using the as.Date function and assign it to a new variable stormDate

```{r}
readStormData$BGN_DATE <- as.Date(readStormData$BGN_DATE, format = "%m%d%Y %H:%m:%s")
class(readStormData$BGN_DATE)
```

Getting the events type as a Data Frame

```{r}
# subsetting the Storm Data
readStormData <- subset(readStormData, 
                        select = c(EVTYPE, FATALITIES, 
                          INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, 
                          CROPDMGEXP))

```


#### 1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?


Since we have already subset the original data based on the EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG and CROPDMGEXP we now need to process the data further in such a way that for each "EVTYPE" we need to find the FATALTIES and INJURIES.

Doing the above process would give us an insight as to which event type caused maximum fatalities and injuries.

```{r}

library(dplyr)

# Aggregating and arranging the Fatalities and Injuries

stormDataFatalities <- arrange(
  aggregate(FATALITIES ~ EVTYPE, data = readStormData, sum),
  desc(FATALITIES), EVTYPE)[1:10,]

# Aggregated data of the Storm Fatalities based on the event type
stormDataFatalities

stormDataInjuries <- arrange(
  aggregate(INJURIES ~ EVTYPE, data = readStormData, sum),
  desc(INJURIES), EVTYPE)[1:10,] 

# Aggregated data of the Storm Injuries based on the event type
stormDataInjuries

```

From both the "stormDataFatalities" and "stormDataInjuries" we can see that event type "TORNADO" has registered the highest number of Fatalities and Injuries, now let is plot the same on the graph.

```{r}
library(lattice)
# plotting the graphs for the Fatalities and Injuries

par(mfrow=c(1,2),mar=c(10,3,3,2))

# Fatalities by event type

barplot(stormDataFatalities$FATALITIES,
        names.arg=stormDataFatalities$EVTYPE,
        las=2,
        col="#FF6504",
        ylab="Fatalities",
        main="Top 10 fatalities by weather event")

# Injuries by event type
barplot(stormDataInjuries$INJURIES,
        names.arg=stormDataInjuries$EVTYPE,
        las=2,
        col="#FF6504",
        ylab="Injuries",
        main="Top 10 Injuries by weather event")

```

#### Across the United States, which types of events have the greatest economic consequences?

The greatest economic consequences can be measured by the columns PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP.

The columns names denote the following

* PROPDMG --> Property Damage
* CROPGMG --> Crop Damage
* PROPDMGEXP --> Property Damage Exponent
* CROPDMGEXP --> Crop Damage Exponent

We need to first associate the Damage caused to the Event type.

To do that we need to convert the notations "K","M","","B","m","+","0","5","6","?","4","2","3","h","7","H","-" "1","8" to their corresponding powers of 10 or exponential values.

We do that by the following operation

```{r}

unique(readStormData$PROPDMGEXP)

# convert the notations "K","M","","B","m","+","0","5","6","?","4","2","3","h","7","H","-" "1","8" to their corresponding powers of 10 or exponential values.

readStormData$PROPEXP[readStormData$PROPDMGEXP == "K"] <- 1000
readStormData$PROPEXP[readStormData$PROPDMGEXP == "M"] <- 1000000
readStormData$PROPEXP[readStormData$PROPDMGEXP == ""] <- 1
readStormData$PROPEXP[readStormData$PROPDMGEXP == "B"] <- 1000000000
readStormData$PROPEXP[readStormData$PROPDMGEXP == "m"] <- 1000000
readStormData$PROPEXP[readStormData$PROPDMGEXP == "0"] <- 1
readStormData$PROPEXP[readStormData$PROPDMGEXP == "5"] <- 100000
readStormData$PROPEXP[readStormData$PROPDMGEXP == "6"] <- 1000000
readStormData$PROPEXP[readStormData$PROPDMGEXP == "4"] <- 10000
readStormData$PROPEXP[readStormData$PROPDMGEXP == "2"] <- 100
readStormData$PROPEXP[readStormData$PROPDMGEXP == "3"] <- 1000
readStormData$PROPEXP[readStormData$PROPDMGEXP == "h"] <- 100
readStormData$PROPEXP[readStormData$PROPDMGEXP == "7"] <- 10000000
readStormData$PROPEXP[readStormData$PROPDMGEXP == "H"] <- 100
readStormData$PROPEXP[readStormData$PROPDMGEXP == "1"] <- 10
readStormData$PROPEXP[readStormData$PROPDMGEXP == "8"] <- 100000000
# Assigning '0' to invalid exponent data
readStormData$PROPEXP[readStormData$PROPDMGEXP == "+"] <- 0
readStormData$PROPEXP[readStormData$PROPDMGEXP == "-"] <- 0
readStormData$PROPEXP[readStormData$PROPDMGEXP == "?"] <- 0

class(readStormData$PROPEXP)
# Calculating the property damage value

stormPropertyDamage <- readStormData$PROPDMG * readStormData$PROPEXP

```

After having converted the notations for property damage we now need to do the same for the crop damage which will be achieved by the following code

```{r}
unique(readStormData$CROPDMGEXP)

# Assigning values for the crop exponent data 
readStormData$CROPEXP[readStormData$CROPDMGEXP == "M"] <- 1000000
readStormData$CROPEXP[readStormData$CROPDMGEXP == "K"] <- 1000
readStormData$CROPEXP[readStormData$CROPDMGEXP == "m"] <- 1000000
readStormData$CROPEXP[readStormData$CROPDMGEXP == "B"] <- 1000000000
readStormData$CROPEXP[readStormData$CROPDMGEXP == "0"] <- 1
readStormData$CROPEXP[readStormData$CROPDMGEXP == "k"] <- 1000
readStormData$CROPEXP[readStormData$CROPDMGEXP == "2"] <- 100
readStormData$CROPEXP[readStormData$CROPDMGEXP == ""] <- 1
# Assigning '0' to invalid exponent data
readStormData$CROPEXP[readStormData$CROPDMGEXP == "?"] <- 0

stormCropDamage <- readStormData$CROPDMG * readStormData$CROPEXP
```